---
title: "framingham"
author: "Sarah Lotspeich"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load packages
library(riskCommunicator) ## for the Framingham teaching data
library(tableone) ## for Table 1
library(ggplot2) ## for pretty plots
library(speedyCMI) ## for imputation methods
```

## Data

```{r}
# Load the data
data(framingham)
length(unique(framingham$RANDID)) ## 4434 patients
```

There are `r length(unique(framingham$RANDID))` patients in the Framingham teaching dataset.

```{r}
to_exclude = framingham |> 
  dplyr::group_by(RANDID) |> 
  dplyr::filter(AGE == min(AGE), ## subset to first visits,
                PREVHYP == 1) |> ## and patients with hypertension at first visit
  dplyr::pull(RANDID) ## create a vector of their IDs
length(to_exclude) # 1430 patients excluded for prevalent hypertension
```

We excluded `r length(to_exclude)` patients who had prevalent hypertension at their first visit in the Framingham Study. We then subsetted the data for the remaining patients to include only visits leading up to and including a hypertension diagnosis (or end of followup).

```{r}
to_analyze = framingham |> 
  dplyr::filter(!(RANDID %in% to_exclude), ## exclude patients with prevalence hypertension (at baseline)
                TIME <= TIMEHYP) ## subset to visits up to and including hypertension diagnosis
```

The analysis data included each patient's first and last visits (either hypertension diagnosis or end of followup), keeping their `SYSBP`, `SEX`, `AGE`, `BMI`, and `TIMEHYP` from each. 

```{r}
to_analyze_first = to_analyze |> 
  dplyr::group_by(RANDID) |> 
  dplyr::filter(AGE == min(AGE)) |> 
  dplyr::rename(SYSBP0 = SYSBP, 
                AGE0 = AGE, 
                BMI0 = BMI) |> 
  dplyr::select(RANDID, SEX, AGE0, BMI0, SYSBP0, TIMEHYP) |> 
  dplyr::mutate(TIMEHYP = TIMEHYP / 365, ## convert time from days --> years
                SEX = factor(x = SEX, 
                             levels = c(1, 2), 
                             labels = c("MALE", "FEMALE"))
                ) 

to_analyze_last = to_analyze |> 
  dplyr::group_by(RANDID) |> 
  dplyr::filter(AGE == max(AGE)) |> 
  dplyr::rename(SYSBP1 = SYSBP, 
                AGE1 = AGE, 
                BMI1 = BMI) |> 
  dplyr::select(RANDID, AGE1, BMI1, SYSBP1, HYPERTEN)

to_analyze = to_analyze_first |> 
  dplyr::left_join(to_analyze_last) |> 
  dplyr::mutate(SYSBP_CHANGE = SYSBP0 - SYSBP1)
```

We excluded `r with(to_analyze, sum(AGE0 == AGE1))` patients who had only one visit before hypertension diagnosis or end of followup. 

```{r}
to_analyze = to_analyze |> 
  dplyr::filter(AGE0 != AGE1) ## subset to patients with at least two visits
nrow(to_analyze) # 2225 patients 
```

Finally, we excluded `r with(to_analyze, sum(is.na(BMI0)))` patients who were missing `BMI` at their first visit. 

```{r}
to_analyze = to_analyze |> 
  dplyr::filter(!is.na(BMI0)) ## subset to patients with non-missing BMI
nrow(to_analyze) # 2218 patients 
```

These exclusions left a sample of `r nrow(to_analyze)` patients for the analysis of change in systolic blood pressure. 

## Exploratory Data Analysis

Patients with and without hypertension are described below. From this table, we observe the following: 

  -   There was a higher proportion of women with hypertension, but a lower proportion of men. 
  -   On average, patients diagnosed with hypertension tended to be older (by about $2$ years) than those who were not diagnosed. 
  -   On average, patients diagnosed with hypertension tended to have higher BMIs (by about $1$ point) than those who were not diagnosed. 
  -   On average, patients diagnosed with hypertension tended to have higher systolic blood pressures at both their first and last visits than those who were not diagnosed.

```{r, echo = FALSE}
tab1 =  CreateTableOne(vars = c("SEX", "AGE0", "BMI0", "SYSBP0", "SYSBP1"), 
                       strata = "HYPERTEN", 
                       data = to_analyze)
print(tab1, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
```

The relationship between systolic blood pressure at first and last visits was positive among patients diagnosed with hypertension or undiagnosed. However, patients diagnosed with hypertension consistently had higher followup systolic blood pressure than patients not diagnosed across all baseline values.

```{r}
to_analyze |> 
  ggplot(aes(x = SYSBP0, y = SYSBP1, color = factor(HYPERTEN))) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm") + 
  theme_minimal()
```

## Imputation Model Selection

```{r}
### Function to make density plot of the distribution of TIMEHYP by diagnosis
density_plot = function(imp_fit, title) {
  imp_fit$imputed_data |> 
    dplyr::mutate(HYPERTEN = factor(x = HYPERTEN, 
                                    levels = c(1, 0), 
                                    labels = c("Diagnosed\n(Uncensored)", "Undiagnosed\n(Censored)"))) |> 
    ggplot(aes(x = imp, fill = HYPERTEN)) + 
    geom_density(alpha = 0.7) + 
    geom_vline(xintercept = 25, linetype = 2) + 
    ggthemes::scale_fill_colorblind(name = "Patient\nStatus:") +
    theme_minimal() + 
    theme(legend.position = "right") + 
    xlab("Time to Hypertension Diagnosis\n(Observed or Imputed)") + 
    ylab("Density") + 
    ggtitle(title) + 
    xlim(c(0, 160))
}
```

### (i) Weibull Distribution
```{r}
## (1) Weibull
weib_imp = cmi_fp_analytical(imputation_model = Surv(time = TIMEHYP, event = HYPERTEN) ~ SEX + AGE0 + BMI0 + SYSBP0, 
                             dist = "weibull", 
                             data = to_analyze)
### Parameters of the fitted model
weib_params = c(weib_imp$coefficients, weib_imp$scale)
weib_params
### AIC for the fitted model
weib_aic = weib_imp$aic
weib_aic
### BIC for the fitted model
weib_bic = weib_imp$bic
weib_bic
### Density plot of the distribution of TIMEHYP
weib_plot = weib_imp |> 
  density_plot(title = "Weibull Imputation Model")
weib_plot
```

### (ii) Exponential Distribution
```{r}
## (2) Exponential
expo_imp = cmi_fp_analytical(imputation_model = Surv(time = TIMEHYP, event = HYPERTEN) ~ SEX + AGE0 + BMI0 + SYSBP0, 
                             dist = "exponential", 
                             data = to_analyze)
### Parameters of the fitted model
expo_params = c(expo_imp$coefficients, expo_imp$scale)
expo_params
### AIC for the fitted model
expo_aic = expo_imp$aic
expo_aic
### BIC for the fitted model
expo_bic = expo_imp$bic 
expo_bic
### Density plot of the distribution of TIMEHYP
expo_plot = expo_imp |> 
  density_plot(title = "Exponential Imputation Model")
expo_plot
```

### (iii) Log-Normal Distribution
```{r}
## (3) Log-normal
lognorm_imp = cmi_fp_analytical(imputation_model = Surv(time = TIMEHYP, event = HYPERTEN) ~ SEX + AGE0 + BMI0 + SYSBP0, 
                                dist = "lognormal", 
                                data = to_analyze)
### Parameters of the fitted model
lognorm_params = c(lognorm_imp$coefficients, lognorm_imp$scale)
lognorm_params
### AIC for the fitted model
lognorm_aic = lognorm_imp$aic
lognorm_aic
### BIC for the fitted model
lognorm_bic = lognorm_imp$bic
lognorm_bic
### Density plot of the distribution of TIMEHYP
lognorm_plot = lognorm_imp |> 
  density_plot(title = "Log-Normal Imputation Model")
lognorm_plot
```

### (iv) Log-Logistic Distribution
```{r}
## (4) Log-logistic
loglog_imp = cmi_fp_analytical(imputation_model = Surv(time = TIMEHYP, event = HYPERTEN) ~ SEX + AGE0 + BMI0 + SYSBP0, 
                               dist = "loglogistic", 
                               data = to_analyze)
### Parameters of the fitted model
loglog_params = c(loglog_imp$coefficients, loglog_imp$scale)
loglog_params
### AIC for the fitted model
loglog_aic = loglog_imp$aic
loglog_aic
### BIC for the fitted model
loglog_bic = loglog_imp$bic
loglog_bic
### Density plot of the distribution of TIMEHYP
loglog_plot = loglog_imp |> 
  density_plot(title = "Log-Logistic Imputation Model")
loglog_plot
```

### (v) Piecewise Exponential Distribution
```{r}
## (5) Piecewise exponential
to_analyze = to_analyze |> 
  dplyr::mutate(FEMALE = as.numeric(SEX == "FEMALE"))
pwe_imp = cmi_fp_analytical(imputation_model = Surv(time = TIMEHYP, event = HYPERTEN) ~ FEMALE + AGE0 + BMI0 + SYSBP0, 
                            dist = "pwe", 
                            data = to_analyze, 
                            nintervals = 10)
### Parameters of the fitted model
pwe_params = c(pwe_imp$coefficients, pwe_imp$logblhaz)
pwe_params
### AIC for the fitted model
pwe_aic = pwe_imp$aic
pwe_aic
### BIC for the fitted model
pwe_bic = pwe_imp$bic
pwe_bic
### Density plot of the distribution of TIMEHYP
pwe_plot = pwe_imp |> 
  density_plot(title = paste("Piecewise Exponential Imputation Model"))
pwe_plot
```

### Comparing Model Fits

```{r}
data.frame(Distribution = c("Exponential", "Log-Logistic", 
                            "Log-Normal", "Piecewise Exponential", 
                            "Weibull"), 
           AIC = c(expo_aic, loglog_aic, lognorm_aic, pwe_aic, weib_aic), 
           BIC = c(expo_bic, loglog_bic, lognorm_bic, pwe_bic, weib_bic)) |> 
  kableExtra::kable(format = "html") |> 
  kableExtra::kable_styling()
```

### Comparing Imputed Values

```{r, fig.height = 7, fig.width = 7, echo = FALSE}
all_plot = weib_imp$imputed_data |> 
  dplyr::mutate(Model = "Weibull") |> 
  dplyr::bind_rows(
    expo_imp$imputed_data |> 
      dplyr::mutate(Model = "Exponential")
  ) |> 
  dplyr::bind_rows(
    lognorm_imp$imputed_data |> 
      dplyr::mutate(Model = "Log-Normal")
  ) |> 
  dplyr::bind_rows(
    loglog_imp$imputed_data |> 
      dplyr::mutate(Model = "Log-Logistic")
  ) |> 
  dplyr::bind_rows(
    pwe_imp$imputed_data |> 
      dplyr::mutate(Model = "Piecewise Exponential")
  ) |> 
  dplyr::mutate(HYPERTEN = factor(x = HYPERTEN, 
                                  levels = c(1, 0), 
                                  labels = c("Diagnosed (Uncensored)", "Undiagnosed (Censored)"))) |> 
  ggplot(aes(x = imp, fill = HYPERTEN)) + 
  geom_density(alpha = 0.7) + 
  geom_vline(xintercept = 25, linetype = 2) + 
  ggthemes::scale_fill_colorblind(name = "Patient Status:") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top", 
        legend.title = element_text(face = "bold"),
        strip.background = element_rect(fill = "black"),
        strip.text = element_text(color = "white"),
        axis.title = element_text(face = "bold")) + 
  xlab("Time to Hypertension Diagnosis (Observed or Imputed) in Years") + 
  ylab("Density") + 
  facet_wrap(~Model)
all_plot
ggsave(filename = "~/Documents/speedyCMI/figures/fig4-density-imputed.png", 
       plot = all_plot, width = 7, height = 7)
```



## Models 

### Naive Analysis (Time to Hypertension or End of Followup)

Among all `r nrow(to_analyze)` patients, whether or not they were diagnosed with hypertension, the adjusted relationship between their systolic blood pressures at the first and last visits is as follows. Note that time to hypertension **or end of followup** is being used "naively" in place of time to hypertension.

```{r}
naive_fit = lm(SYSBP1 ~ SYSBP0 + TIMEHYP + SEX + AGE0 + BMI0, 
               data = to_analyze)
knitr::kable(coefficients(summary(naive_fit)), digits = 3)
knitr::kable(confint(naive_fit), digits = 3)
forest_dat = coefficients(summary(naive_fit)) |> 
  data.frame() |> 
  dplyr::mutate(Coefficient = rownames(coefficients(summary(naive_fit))),
                Method = "Naive Analysis") |> 
  dplyr::inner_join(
    confint(naive_fit) |> 
      data.frame() |> 
      dplyr::mutate(Coefficient = rownames(confint(naive_fit))) |> 
      magrittr::set_colnames(c("LB", "UB", "Coefficient"))
  )
```

  -   Higher systolic blood pressures at first visit were associated with higher systolic blood pressures at last visit. On average, for every 1 mmHg higher a patient's systolic blood pressure was at first visit, their systolic blood pressure was expected to be `r round(naive_fit$coefficients[2], 3)` mmHg *higher* at last visit.
  -   Patients who were farther from a hypertension diagnosis at first visit (i.e., larger `TIMEHYP`) were expected to have lower systolic blood pressures at last visit, on average. Specifically, for every $1$ year farther away they were from their hypertension diagnosis at their first visit, their systolic blood pressure at that time was expected to be `r abs(round(naive_fit$coefficients[3], 3))` mmHg *lower* at their last visit.
  -   Patients who were older at their first visit had  higher systolic blood pressures at their last visit, on average. For every year older a patient was at their first visit, their systolic blood pressure was expected to be `r round(naive_fit$coefficients[4], 3)` mmHg *higher* at last visit.
  -   Patients with higher BMIs at their first visit hadlower systolic blood pressures at their last visit, on average. For every $1$ point higher a patient's BMI was at their first visit, their systolic blood pressure was expected to be `r abs(round(naive_fit$coefficients[5], 3))` mmHg *lower* at last visit.

### Complete Case Analysis (Time to Hypertension)

Among only the subset of `r sum(to_analyze$HYPERTEN)` patients who were diagnosed with hypertension, the adjusted relationship between their systolic blood pressures at the first and last visits is as follows.

```{r}
cc_fit = lm(SYSBP1 ~ SYSBP0 + TIMEHYP + SEX + AGE0 + BMI0, 
            data = to_analyze, 
            subset = HYPERTEN == 1)
knitr::kable(coefficients(summary(cc_fit)), digits = 3)
knitr::kable(confint(cc_fit), digits = 3)
forest_dat = coefficients(summary(cc_fit)) |> 
  data.frame() |> 
  dplyr::mutate(Coefficient = rownames(coefficients(summary(cc_fit))),
                Method = "Complete Case Analysis") |> 
  dplyr::inner_join(
    confint(cc_fit) |> 
      data.frame() |> 
      dplyr::mutate(Coefficient = rownames(confint(cc_fit))) |> 
      magrittr::set_colnames(c("LB", "UB", "Coefficient"))
  ) |> 
  dplyr::bind_rows(forest_dat)
```

  -   Higher systolic blood pressures at first visit were associated with higher systolic blood pressures at last visit. On average, for every 1 mmHg higher a patient's systolic blood pressure was at first visit, their systolic blood pressure was expected to be `r round(cc_fit$coefficients[2], 3)` mmHg *higher* at last visit.
  -   Patients who were farther from a hypertension diagnosis at first visit (i.e., larger `TIMEHYP`) were expected to have lower systolic blood pressures at last visit, on average. Specifically, for every $1$ year farther away they were from their hypertension diagnosis at their first visit, their systolic blood pressure at that time was expected to be `r abs(round(cc_fit$coefficients[3], 3))` mmHg *lower* at their last visit.
  -   Patients who were older at their first visit had  higher systolic blood pressures at their last visit, on average. For every year older a patient was at their first visit, their systolic blood pressure was expected to be `r round(cc_fit$coefficients[4], 3)` mmHg *higher* at last visit.
  -   Patients with higher BMIs at their first visit hadlower systolic blood pressures at their last visit, on average. For every $1$ point higher a patient's BMI was at their first visit, their systolic blood pressure was expected to be `r abs(round(cc_fit$coefficients[5], 3))` mmHg *lower* at last visit.

### Multiple Imputation Analysis

```{r}
set.seed(333)
## Fit analysis model to each imputed dataset
lognorm_imp = cmi_fp_analytical(imputation_model = Surv(time = TIMEHYP, event = HYPERTEN) ~ SEX + AGE0 + BMI0 + SYSBP0, 
                                dist = "lognormal", 
                                data = to_analyze, 
                                boots = 20)
mult_fit = do.call(what = rbind,
                   args = lapply(X = lognorm_imp,
                                 FUN = function(l) {
                                   summary(lm(formula = SYSBP1 ~ SYSBP0 + imp + SEX + AGE0 + BMI0, data = l$imputed_data))$coefficients
                                 }
                   )
)

## Pool estimates
### Take sums of estimates and variances
sum_est_var = rowsum(x = mult_fit[, 1:2],
                     group = rownames(mult_fit))
### Divide by the number of imputations to get means
mean_est_var = sum_est_var / 20

## Save parameter estimates
mean_est_var
rownames(mean_est_var)[4] = "TIMEHYP"
forest_dat = mean_est_var |>
  data.frame() |> 
  dplyr::mutate(Coefficient = rownames(mean_est_var),
                Method = "Multiple Imputation Analysis", 
                LB = Estimate - 1.96 * Std..Error, 
                UB = Estimate + 1.96 * Std..Error) |>
  dplyr::bind_rows(forest_dat)
```

### Comparing Methods 

```{r, fig.width = 8, fig.height = 6}
forest_plot = forest_dat |> 
  dplyr::mutate(Method = factor(x = Method, 
                                levels = c("Naive Analysis", "Multiple Imputation Analysis", "Complete Case Analysis")),#, 
                                #labels = c("Naive\nAnalysis", "MLE\nAnalysis", "Complete\nCase\nAnalysis")),
                Coefficient = factor(x = Coefficient, 
                                     levels = c("(Intercept)", "SYSBP0", "TIMEHYP", 
                                                "SEXFEMALE", "AGE0", "BMI0"), 
                                     labels = c("INTERCEPT", "SYSBP0", "TIME", 
                                                "FEMALE", "AGE", "BMI"))) |> 
  ggplot(aes(x = Method, y = Estimate, color = Method)) + 
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(position = position_dodge(width = 1), 
             size = 1.5) + 
  geom_errorbar(aes(ymin = LB, ymax = UB), 
                position = position_dodge(width = 1), 
                linewidth = 1) + 
  theme_minimal(base_size = 14) +
  facet_wrap(~Coefficient, scales = "free") + 
  ggthemes::scale_color_colorblind(name = "Method:") + 
  theme(legend.position = "top",
        legend.title = element_text(face = "bold"),
        strip.background = element_rect(fill = "black"),
        strip.text = element_text(color = "white"),
        axis.title = element_text(face = "bold"), 
        axis.text.x = element_blank(), 
        axis.title.x = element_blank())
forest_plot
ggsave(filename = "~/Documents/speedyCMI/figures/fig5-forest-plot.png", 
       plot = forest_plot, width = 8, height = 5)
```

### Comparing Imputation Models 

```{r}
set.seed(114)
forest_dat = data.frame()
for (d in c("weibull", "exponential", "lognormal", "loglogistic")) {
  ## Fit analysis model to each imputed dataset
  imp = cmi_fp_analytical(imputation_model = Surv(time = TIMEHYP, event = HYPERTEN) ~ SEX + AGE0 + BMI0 + SYSBP0, 
                                  dist = "lognormal", 
                                  data = to_analyze, 
                                  boots = 20)
  mult_fit = do.call(what = rbind,
                     args = lapply(X = imp,
                                   FUN = function(l) {
                                     summary(lm(formula = SYSBP1 ~ SYSBP0 + imp + SEX + AGE0 + BMI0, data = l$imputed_data))$coefficients
                                   }
                     )
  )
  
  ## Pool estimates
  ### Take sums of estimates and variances
  sum_est_var = rowsum(x = mult_fit[, 1:2],
                       group = rownames(mult_fit))
  ### Divide by the number of imputations to get means
  mean_est_var = sum_est_var / 20
  
  ## Save parameter estimates
  mean_est_var
  rownames(mean_est_var)[4] = "TIMEHYP"
  forest_dat = mean_est_var |>
    data.frame() |> 
    dplyr::mutate(Coefficient = rownames(mean_est_var),
                  Method = toupper(d), 
                  LB = Estimate - 1.96 * Std..Error, 
                  UB = Estimate + 1.96 * Std..Error) |>
    dplyr::bind_rows(forest_dat)  
}
```

```{r, fig.width = 8, fig.height = 6}
forest_plot = forest_dat |> 
  dplyr::mutate(Coefficient = factor(x = Coefficient, 
                                     levels = c("(Intercept)", "SYSBP0", "TIMEHYP", 
                                                "SEXFEMALE", "AGE0", "BMI0"), 
                                     labels = c("INTERCEPT", "SYSBP0", "TIME", 
                                                "FEMALE", "AGE", "BMI")),
                Method = factor(x = Method, 
                                levels = c("LOGNORMAL", "LOGLOGISTIC", "WEIBULL", "EXPONENTIAL"), 
                                labels = c("Log-Normal", "Log-Logistic", "Weibull", "Exponential"))) |> 
  ggplot(aes(x = Method, y = Estimate, color = Method)) + 
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(position = position_dodge(width = 0.5), 
             size = 1.5) + 
  geom_errorbar(aes(ymin = LB, ymax = UB), 
                position = position_dodge(width = 0.5), 
                linewidth = 1) + 
  theme_minimal(base_size = 14) +
  facet_wrap(~Coefficient, scales = "free") + 
  ggthemes::scale_color_colorblind(name = "Imputation Model:") + 
  theme(legend.position = "top",
        legend.title = element_text(face = "bold"),
        strip.background = element_rect(fill = "black"),
        strip.text = element_text(color = "white"),
        axis.title = element_text(face = "bold"), 
        axis.text.x = element_blank(), 
        axis.title.x = element_blank())
forest_plot
ggsave(filename = "~/Documents/speedyCMI/figures/fig6-forest-plot-imp.png", 
       plot = forest_plot, width = 8, height = 5)
```
